---
title: "Cut Circle: Intonation and Synchronization"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", echo = FALSE, warnings = FALSE, message = FALSE)
options(tidyverse.quite = T)
library(targets)
library(tidyverse)
tar_make()
```

# Method

## Procedure

The choir sung three Rennaissance pieces in three different conditions three times on three different days, except for 
Guillaume Dufay's "Missa Ecce ancilla domini", which was only sung twice due to a recording error on the first day. Each condition and work comprises one take, with 24 takes in total. 

The three experimental conditions were related to the physical setup of singers: "Far Apart, No Touch" (*far-apart*), "Close, No Touch" (*no-touch*), and "Close, Touch" (*touch*). The scores were projected via a projector. See Lange et. al. for more information on the setup details.  

The two works by Josquin consisted of a single parts, whereas the Dufay masses consists of three parts each, which were partly repeated in a take in order to achieve a recording duration of at least 6 minutes for the physiological measurement. The same holds true for Josquin's "D'ung Altre".

All works execpt Dufay's Agnus II were set in four-part harmony (named here consistently as soprano, alto, tenor, and bass). Dufay's Agnus II features  only soprano and alto. All voices were sung by two singers, singing in unisone except for a few final notes, which were split in two (dufay-agnus1, dufay-kyrie1) or on a single occasion in three tones (dufay-kyrie1). 

All singers were equipped with headsets and recorded (soprano: headsets 7 and 8, alto: headsets 1 and 2, tenor: headsets 5 and 6, bass: headsets 3 and 4). The recordings were transcribed with the help of Tony software (REF TONY) and manually corrected and optimized where the algorithm gave erroneous or sub-optimal results.
Subsequently, the note tracks with exact onsets,duration, and pitches (measured in Hz) were imported into Sonic Visualiser and in manually annotated with note labels, which were generated from digital scores of all works with the help of **music21**. During this step, extra and misplaced notes were also annotated by an musicologist. All annotated note tracks were exported in CSV format and imported into R for further analysis, along with various metadata. This resulted in 672 note tracks with a total of `r nrow(tar_read(note_tracks))` tone events.

## Measures
In 


```{r piece_stats}
tar_read(piece_info) %>% 
  select(ID = piece, Work = full_name) %>% 
  flextable::flextable() %>% 
  flextable::set_caption("Works used in the study") %>% 
  flextable::autofit()
```


```{r take_stats, echo = FALSE}
tar_read(clean_note_tracks) %>% 
  #>distinct(piece, take, condition, headset) %>% 
  group_by(piece, take, condition, repetition) %>% 
  summarise(number_of_notes = n(), no_voices = n_distinct(voice_type), .groups = "drop") %>%
  select(Take = take, 
         Piece = piece, 
         Repetition = repetition, 
         Condition = condition, 
         `No. Notes`= number_of_notes,
         `No. Voices` = no_voices) %>% 
  arrange(Take, Repetition, Piece) %>% 
  flextable::flextable() %>% 
  flextable::set_caption("Basic statistics of the single takes.") %>% 
  flextable::autofit()
```

